### 2018-02-10 submission/manuscript.Rmd
update changes in intro

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying microbial signatures of tumors (biomarkers) but did so on a small total number of individuals and only investigated stool. This present meta-analysis addresses some of these major shortcomings.

Although there has been an intense focus on microbiota-based biomarker discovery for tumors, the number of candidate genera seem to be endless. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and colorectal tumors, making it hard to know how extrapolatable these findings are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of disease that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015]. Additionally, genera like *Roseburia* have been found in some studies to be increased in tumors but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera of disease. Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of disease versus controls. Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. Based on the studies mentioned, there seems to be very little overlap between the genera identified to be associated with adenoma and carcinoma, with *Lactobacillus* being one of the few commonalities.

Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without limitations. The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of colorectal tumors. Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased odds ratio (OR) for adenoma or carcinoma stages of disease We then assessed what genera, if any, increase or decrease the OR of adenoma or carcinoma stages of disease Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased OR genera resulted in better model classification, based on the area under the curve (AUC). Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power. Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.

@@

***Study Analysis Overview:***
OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and OR. For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in OR for having an adenoma or carcinoma and ranked based on P-value. We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased OR based on P-value). Comparison between the full and selected models took two different approaches. In the first approach, models were trained on one study then tested on the remaining studies. This process was repeated for every study in the meta-analysis. In the second approach models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which only control versus adenoma indivdiuals were made, the carcinoma individuals were excluded from each respective study. Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study. For all analysis completed feces and tissue samples were kept seperate. Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).

***Obtaining Genera Relative Abundance and Selected Models:***
For the genera analysis of the OR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings. The select models for the first approach utilized the top 5 most significantly increased and decreased ORs. For this approach, the full community models ufor each study were built by utilizing all genera and lowest taxnomic groups identified within that particular study. The select models for the second approach, that used OTU-based Random Forest models, utilized a similar method as the first approach. The main difference was that any OTU that taxonomically classified to one of the genera in the top 5 increased or decreased OR were included in the select model. OTU Random Forest models using the full community included all OTUs.

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case-control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying microbial signatures of tumors (biomarkers) but did so on a small total number of individuals and only investigated stool. This present meta-analysis addresses some of these major shortcomings.

Although there has been an intense focus on microbiota-based biomarker discovery for tumors, the number of candidate genera seem to be endless. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with tumors. These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, the sample size (n = 509) is equal to or less than some of the more recent individual studies investigating the microbiota and colorectal tumors, making it hard to know how extrapolatable these findings are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of disease that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015]. Additionally, genera like *Roseburia* have been found in some studies to be increased in tumors but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on individuals with carcinomas but associations with the adenoma stage of disease are not any clearer at identifying candidate genera correlated with these earlier tumors. Groups focusing on broad scale community metrics have found that measures such as richness are decreased in those with adenomas versus controls. Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in those with adenoma tumors [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. Additionally, based on these studies mentioned, there seems to be very few common genera that are associated with both adenoma and carcinoma tumors, with *Lactobacillus* being one of the few commonalities.

Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk of disease. The current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without their own limitations. After the colonoscopy bowel prep the bacterial community sampled may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of colorectal tumors. Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in individuals with adenoma or carcinoma and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased odds ratio (OR) for individuals to have either an adenoma or carcinoma. We then assessed what genera, if any, increase or decrease the OR of an individual to have an adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased OR genera resulted in better model classification, based on the area under the curve (AUC). Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power. Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the individual studies are underpowered for assessing small effect sizes.

@@

***Study Analysis Overview:***
OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and OR. For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in OR for having an adenoma or carcinoma and ranked based on P-value. We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased OR based on P-value). Comparison between the full and selected models took two different approaches. In the first approach, models were trained on one study then tested on the remaining studies. This process was repeated for every study in the meta-analysis. In the second approach models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which only control versus adenoma individuals were made, the carcinoma individuals were excluded from each respective study. Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study. For all analysis completed feces and tissue samples were kept separate. Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).

***Obtaining Genera Relative Abundance and Selected Models:***
For the genera analysis of the OR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings. The select models for the first approach utilized the top 5 most significantly increased and decreased ORs. For this approach, the full community models for each study were built by utilizing all genera and lowest taxonomic groups identified within that particular study. The select models for the second approach, that used OTU-based Random Forest models, utilized a similar method as the first approach. The main difference was that any OTU that taxonomically classified to one of the genera in the top 5 increased or decreased OR were included in the select model. OTU Random Forest models using the full community included all OTUs.

### 2018-02-10 submission/manuscript.tex
update changes in intro

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been purported to play an active role in disease
pathogenesis {[}1,2{]}. Numerous studies have shown the importance of
both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case/control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis focused
on identifying microbial signatures of tumors (biomarkers) but did so on
a small total number of individuals and only investigated stool. This
present meta-analysis addresses some of these major shortcomings.

Although there has been an intense focus on microbiota-based biomarker
discovery for tumors, the number of candidate genera seem to be endless.
Some studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
these studies, mouth-associated genera are far from the only microbes
identified to be associated with CRC. These other genera include, but
are not limited to, \emph{Providencia}, \emph{Mogibacterium},
\emph{Enterococcus}, \emph{Escherichia/Shigella}, \emph{Klebsiella}, and
\emph{Streptococcus} {[}14--16{]}. In fact, there is good \emph{in vivo}
evidence that \emph{Escherichia/Shigella} and \emph{Streptococcus} can
be important in the pathogenesis of CRC {[}5,19,20{]}. Other studies
have also identified \emph{Akkermansia muciniphila} and
\emph{Bacteroides fragilis} as potential markers of CRC with good
mechanistic studies for the latter {[}15,21,22{]}. A recent
meta-analysis confirmed the correlations of certain mouth-associated
genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
However, the sample size (n = 509) is equal to or less than some more
recent individual studies investigating the microbiota and colorectal
tumors, making it hard to know how extrapolatable these findings are.
That particular meta-analysis also added more potential microbial
associations to both carcinoma (\emph{Pantoea agglomerans}
\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
\emph{Lactobacillus}) stages of disease that need to be investigated
further, since a number of these genera have been found to be enriched
in controls and not disease {[}13,16,17{]}. Additionally, genera like
\emph{Roseburia} have been found in some studies to be increased in
tumors but in others to either be decreased or have no difference
{[}15,18,23,24{]}.

Most of these studies have focused on carcinoma but the adenoma
observations are not any clearer at identifying candidate genera of
disease. Groups focusing on broad scale community metrics have found
that metrics such as richness are decreased in the adenoma stage of
disease versus controls. Other studies have identified
\emph{Lactococcus}, \emph{Pseudomonas}, \emph{Acidovorax},
\emph{Cloacibacterium}, \emph{Helicobacter}, \emph{Lactobacillus},
\emph{Bilophila}, \emph{Desulfovibrio}, and \emph{Mogibacterium} to be
increased in adenoma {[}25--27{]}. Based on the studies mentioned, there
seems to be very little overlap between the genera identified to be
associated with adenoma and carcinoma, with \emph{Lactobacillus} being
one of the few commonalities.

Targeting the identification of tumor microbial biomarkers within stool
seems logical since it offers an easy and cost-effective way to stratify
risk and the current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not without risk of complications. Although stool
represents an easy and less invasive way to assess risk, it is not clear
how well this sample reflects adenoma- and carcinoma- associated
microbial communities. Some studies have tried to assess this in health
and disease but are limited by their sample size {[}18,28{]}. Sampling
the microbiota directly associated with colon tissue may provide clearer
answers but is not without limitations. The community present for
sampling following the colonoscopy bowel prep may reflect the better
adhered microbiota versus the resident community. Additionally, these
samples contain more host DNA, potentially limiting the types of
analysis that can be done. It is well known that low biomass samples can
be very difficult to work with and results can be study dependent due to
the randomness of contamination {[}29{]}.

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, re-examines important
genera across adenoma and carcinoma across study, and examines
differences and similarities between stool and tissue microbiota in the
context of colorectal tumors. Importantly, this analysis and approach
could provide valuable insights into the common genera that are both
protective and detrimental in CRC and whether broad bacterial community
measurements can account for these changes that were not provided by
earlier meta-analysis studies {[}11{]}.

Using both feces (n = 1737) and colon tissues (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community or specific members are more important risk factors
for both adenoma and carcinoma stages of disease. To accomplish this we
first assessed whether bacterial diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased odds
ratio (OR) for adenoma or carcinoma stages of disease We then assessed
what genera, if any, increase or decrease the OR of adenoma or carcinoma
stages of disease Next, using Random Forest models, we analyzed whether
the full community or only the combined top 5 increased and top 5
decreased OR genera resulted in better model classification, based on
the area under the curve (AUC). Finally, we also examined at what effect
and sample size the studies used were powered for and the sample size
needed to get to the traditionally accepted 80\% power. Our results from
these analyses suggests that the bacterial community changes as disease
severity worsens, that specific members are important for disease
classification, and that many of the studies are underpowered for
assessing small effect sizes.

@@

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity was first assessed for differences between controls,
adenoma stage, and carcinoma stage using both linear mixed-effect models
and OR. For each individual study the Bray-Curtis index was used to
assess differences between control-adenoma and control-carcinoma. Next,
all common genera were assessed for differences in OR for having an
adenoma or carcinoma and ranked based on P-value. We then built Random
Forest models based on the full or selected community (the top 5
increased and top 5 decreased OR based on P-value). Comparison between
the full and selected models took two different approaches. In the first
approach, models were trained on one study then tested on the remaining
studies. This process was repeated for every study in the meta-analysis.
In the second approach models were built using OTU level data and a
10-fold CV over 100 different iterations, based on random 80/20
splitting of the data, was used to generate a range of expected AUCs.
For these OTU-based models, the selected model included all OTUs that
had a taxonomic classification to a taxa in the top 5 increased and top
5 decreased OR based on P-value. Finally, the power of each study was
assessed for an effect size ranging from 1\% to 30\% and an estimated
sample size, for these effect sizes, was generated based on 80\% power.
For comparisons in which only control versus adenoma indivdiuals were
made, the carcinoma individuals were excluded from each respective
study. Similarly, for comparisons in which control versus carcinoma
individuals were made the adenoma individuals were excluded from each
respective study. For all analysis completed feces and tissue samples
were kept seperate. Within the tissue groups the data were further
divided between samples from the same individual (matched) and those
from different individuals (unmatched).

\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
For the genera analysis of the OR, OTUs were added together based on the
genus or lowest available taxonomic classification level and the total
average counts, for 100 different subsamplings. The select models for
the first approach utilized the top 5 most significantly increased and
decreased ORs. For this approach, the full community models ufor each
study were built by utilizing all genera and lowest taxnomic groups
identified within that particular study. The select models for the
second approach, that used OTU-based Random Forest models, utilized a
similar method as the first approach. The main difference was that any
OTU that taxonomically classified to one of the genera in the top 5
increased or decreased OR were included in the select model. OTU Random
Forest models using the full community included all OTUs.

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been purported to play an active role in disease
pathogenesis {[}1,2{]}. Numerous studies have shown the importance of
both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case-control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis focused
on identifying microbial signatures of tumors (biomarkers) but did so on
a small total number of individuals and only investigated stool. This
present meta-analysis addresses some of these major shortcomings.

Although there has been an intense focus on microbiota-based biomarker
discovery for tumors, the number of candidate genera seem to be endless.
Some studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
these studies, mouth-associated genera are far from the only microbes
identified to be associated with tumors. These other genera include, but
are not limited to, \emph{Providencia}, \emph{Mogibacterium},
\emph{Enterococcus}, \emph{Escherichia/Shigella}, \emph{Klebsiella}, and
\emph{Streptococcus} {[}14--16{]}. In fact, there is good \emph{in vivo}
evidence that \emph{Escherichia/Shigella} and \emph{Streptococcus} can
be important in the pathogenesis of CRC {[}5,19,20{]}. Other studies
have also identified \emph{Akkermansia muciniphila} and
\emph{Bacteroides fragilis} as potential markers of CRC with good
mechanistic studies for the latter {[}15,21,22{]}. A recent
meta-analysis confirmed the correlations of certain mouth-associated
genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
However, the sample size (n = 509) is equal to or less than some of the
more recent individual studies investigating the microbiota and
colorectal tumors, making it hard to know how extrapolatable these
findings are. That particular meta-analysis also added more potential
microbial associations to both carcinoma (\emph{Pantoea agglomerans}
\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
\emph{Lactobacillus}) stages of disease that need to be investigated
further, since a number of these genera have been found to be enriched
in controls and not disease {[}13,16,17{]}. Additionally, genera like
\emph{Roseburia} have been found in some studies to be increased in
tumors but in others to either be decreased or have no difference
{[}15,18,23,24{]}.

Most of these studies have focused on individuals with carcinomas but
associations with the adenoma stage of disease are not any clearer at
identifying candidate genera correlated with these earlier tumors.
Groups focusing on broad scale community metrics have found that
measures such as richness are decreased in those with adenomas versus
controls. Other studies have identified \emph{Lactococcus},
\emph{Pseudomonas}, \emph{Acidovorax}, \emph{Cloacibacterium},
\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Bilophila},
\emph{Desulfovibrio}, and \emph{Mogibacterium} to be increased in those
with adenoma tumors {[}25--27{]}. Additionally, based on these studies
mentioned, there seems to be very few common genera that are associated
with both adenoma and carcinoma tumors, with \emph{Lactobacillus} being
one of the few commonalities.

Targeting the identification of tumor microbial biomarkers within stool
seems logical since it offers an easy and cost-effective way to stratify
risk of disease. The current gold standard for diagnosis, a colonoscopy,
can be time-consuming and is not without risk of complications. Although
stool represents an easy and less invasive way to assess risk, it is not
clear how well this sample reflects adenoma- and carcinoma- associated
microbial communities. Some studies have tried to assess this in health
and disease but are limited by their sample size {[}18,28{]}. Sampling
the microbiota directly associated with colon tissue may provide clearer
answers but is not without their own limitations. After the colonoscopy
bowel prep the bacterial community sampled may reflect the better
adhered microbiota versus the resident community. Additionally, these
samples contain more host DNA, potentially limiting the types of
analysis that can be done. It is well known that low biomass samples can
be very difficult to work with and results can be study dependent due to
the randomness of contamination {[}29{]}.

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, re-examines important
genera across adenoma and carcinoma across study, and examines
differences and similarities between stool and tissue microbiota in the
context of colorectal tumors. Importantly, this analysis and approach
could provide valuable insights into the common genera that are both
protective and detrimental in individuals with adenoma or carcinoma and
whether broad bacterial community measurements can account for these
changes that were not provided by earlier meta-analysis studies
{[}11{]}.

Using both feces (n = 1737) and colon tissues (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community or specific members are more important risk factors
for both adenoma and carcinoma stages of disease. To accomplish this we
first assessed whether bacterial diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased odds
ratio (OR) for individuals to have either an adenoma or carcinoma. We
then assessed what genera, if any, increase or decrease the OR of an
individual to have an adenoma or carcinoma. Next, using Random Forest
models, we analyzed whether the full community or only the combined top
5 increased and top 5 decreased OR genera resulted in better model
classification, based on the area under the curve (AUC). Finally, we
also examined at what effect and sample size the studies used were
powered for and the sample size needed to get to the traditionally
accepted 80\% power. Our results from these analyses suggests that the
bacterial community changes as disease severity worsens, that specific
members are important for disease classification, and that many of the
individual studies are underpowered for assessing small effect sizes.

@@

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity was first assessed for differences between controls,
adenoma stage, and carcinoma stage using both linear mixed-effect models
and OR. For each individual study the Bray-Curtis index was used to
assess differences between control-adenoma and control-carcinoma. Next,
all common genera were assessed for differences in OR for having an
adenoma or carcinoma and ranked based on P-value. We then built Random
Forest models based on the full or selected community (the top 5
increased and top 5 decreased OR based on P-value). Comparison between
the full and selected models took two different approaches. In the first
approach, models were trained on one study then tested on the remaining
studies. This process was repeated for every study in the meta-analysis.
In the second approach models were built using OTU level data and a
10-fold CV over 100 different iterations, based on random 80/20
splitting of the data, was used to generate a range of expected AUCs.
For these OTU-based models, the selected model included all OTUs that
had a taxonomic classification to a taxa in the top 5 increased and top
5 decreased OR based on P-value. Finally, the power of each study was
assessed for an effect size ranging from 1\% to 30\% and an estimated
sample size, for these effect sizes, was generated based on 80\% power.
For comparisons in which only control versus adenoma individuals were
made, the carcinoma individuals were excluded from each respective
study. Similarly, for comparisons in which control versus carcinoma
individuals were made the adenoma individuals were excluded from each
respective study. For all analysis completed feces and tissue samples
were kept separate. Within the tissue groups the data were further
divided between samples from the same individual (matched) and those
from different individuals (unmatched).

\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
For the genera analysis of the OR, OTUs were added together based on the
genus or lowest available taxonomic classification level and the total
average counts, for 100 different subsamplings. The select models for
the first approach utilized the top 5 most significantly increased and
decreased ORs. For this approach, the full community models for each
study were built by utilizing all genera and lowest taxonomic groups
identified within that particular study. The select models for the
second approach, that used OTU-based Random Forest models, utilized a
similar method as the first approach. The main difference was that any
OTU that taxonomically classified to one of the genera in the top 5
increased or decreased OR were included in the select model. OTU Random
Forest models using the full community included all OTUs.
